What We Eval in the Shadows

### 2021-04-15 paper/main.tex
Apply ck's proofread in 5.4

The code executed by \eval can do side effects. From the compiler perspective,
we care about the observable side effects in environments---\ie variable
definitions, updates, and removals that are visible after a call to \eval
finishes. Knowing where---\ie in which environments do these side-effects
happen, can help us to determine how much of the compiler knowledge about the
program will be potentially invalidated.

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. In the tracer, we register callbacks
that record each environment creation and removal together with updates to its
bindings. From the recorded data, we ignore side effects
coming from \eval sites from unit testing frameworks.\footnote{In the corpus we
have \c{RUnit, testthat, tinytest} and \c{unitizer} unit testing frameworks.}
They run the testing code via \eval and thus the results would be biased
because of the high number of code run in these tests
(\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
the accidental side-effects that are caused by the R virtual machine
implementation and are not related to the user code. For example, the
\c{.Random.seed} variable, which contains the state of the random number
generator is saved and restored from and to the global environment every time
user calls one of the base routines for getting random numbers. Removing them
leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). These
sites are part of \SEUserFunctions in \SEUserPackages packages. \SEFunsNighty
functions are responsible for 90\% of side effects. Half of the side-effects
comes just from three functions: \c{plyr::allocate\_column} (allocates space
for a new data frame column), \c{withr::execute\_handlers} (executes deferred
expressions) and \c{foreach::doSEQ} (executes an expression on each element in
a collection, possibly in parallel).

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environment and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object} which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class means that all of its \eval sites that
do side effect must have the same target environment class. Majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites and over a third of the functions do side effects in either \emph{Local}
or \emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

@@

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls we see mostly assignments and in terms of
sites definitions. This is expected. A subsequent \eval call will turn a
definition into an update. Variable removal, while rare, it does happen.
However, the vast majority comes from the already mentioned
\c{withr::execute\_handlers} function used to defer evaluation of an expression
to after the function exit. In our dataset, it is used almost exclusively by
the \c{tidyselect} package that uses it to remove the reference to current
quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

>>>>

The code executed by \eval can do side effects. From the compiler perspective,
we care about the observable side effects in environments---\ie variable
definitions, updates, and removals that are visible after a call to \eval
finishes. Knowing where---\ie in which environments do these side-effects
happen, can help us determine how much of the compiler knowledge about the
program will be potentially invalidated.

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. We register callbacks in the tracer
that records each environment creation and removal with updates to its
bindings. From the recorded data, we ignore side effects coming from \eval
sites from unit testing frameworks.\footnote{In the corpus we have \c{RUnit,
testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
testing code via \eval, and thus the results would be biassed because of the
high number of code in these tests (\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
the accidental side-effects caused by the R virtual machine implementation and
are not related to the user code. For example, the \c{.Random.seed} variable,
which contains the state of the random number generator is saved and restored
from and to the global environment every time the user calls one of the base
routines that use random numbers. Removing them leaves us with \SEUser side
effects (\SEUserRatio) from \SEUserCalls \eval calls (\SEUserCallsRatio) in
\SEUserSites sites (\SEUserSitesRatio). These sites are part of
\SEUserFunctions in \SEUserPackages packages. \SEFunsNighty functions are
responsible for 90\% of side effects. Half of the side-effects come just from
three functions: \c{plyr::allocate\_column} (allocates space for a new data
frame column), \c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel).

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class, all of its \eval sites that do side
effects must have the same target environment class The majority of the \eval
sites (\SESitesInOneClass) do all side effects consistently in one environment
class. The same happens on the function level. Almost half of the sites (over a
third of the functions) do side effects in either \emph{Local} or \emph{Object}
environments. This gives a ray of hope for the compiler. Even though it is
possible to do anything anywhere, the data suggest that most side-effects are
sane.

@@

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls, we see primarily assignments and in
terms of site definitions. This is expected. A subsequent \eval call will turn
a definition into an update. The most dangerous side effect is variable removal
as it means that after a call to \eval some binding in some environments will
disappear. While rare, this happens, but the vast majority comes from a single
site called by the already mentioned \c{withr::execute\_handlers} function.
This makes sense as it is used to defer evaluation of an expression to after
the function exit and thus used for clean up. In the corpus, it is used almost
exclusively by the \c{tidyselect} package for removing the reference to the
current quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

% TODO type of variables
% TODO do assignments change types?
